Leveraging genomics to detect trait selection from within and between-species variations

### 2023-09-19 manuscript/main.tex
Include N. Salamin comments

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

[not sure of this paragraph...]

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
Additionally, contrasting both trait and genetic differentiation across several populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of gene expression in different populations~\parencite{whitehead_neutral_2006, gilad_natural_2006}. [why only example from gene expression?]
However, dominant selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.

@@

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value~\parencite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
However, in the context of detecting selection acting on a trait, within-species variation is mostly used to describe technical errors.
One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
For population geneticists, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the mean of variances

@@

The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ beinf a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the mean phenotypic trait value changes along the phylogeny, allowing estimating of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within-species, the genetic variance allows estimating of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

\subsubsection*{Bayesian estimate}
[NS: the start of this paragraph is not so clear -> to discuss]
Even though the neutrality index can be applied to any trait, we assume here that the variation between several traits is independent.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ non-independent traits evolving along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is Gibbs sampled.
For each trait and each species, prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait is combined with nucleotide diversity to compute $\EstRateWhithin$ for each species, and is averaged out across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

@@

% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
% pS = 0.0027577478571428568
% Tree length (dS) = 2.340579
% Tree length (My) = 1259.4490200000002
% Root age (My) = 98.9489413888889
% Root age (dS) = 0.18388820079995308
% nbr_sites_var = 27.577478571428568
% u = 1.3788739285714283e-05
% nbr_generations = 13336.114128321291
We tested the xxx [NS: explain a bit what we test here -> NI]
We simulated phenotypic values as the sum of genotypic value and an environmental effect.
The environmental effect was normally distributed with variance $\VarEnv$.
We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
We followed a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
We scaled parameters to fit plausible values for mammals.
We used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.

The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
Additionally, the instant value of log-$\Ne$ was modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
The geometric Brownian motion accounted for long-term fluctuations, while the Ornstein-Uhlenbeck introduced short-term fluctuations. [NS: not clear how you combined both here]
The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.

@@

For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
At the species level, the additive genetic variance of the neutral trait ($\VarGenetic$) is normalized by nucleotide diversity of any neutral genomic region ($\pi$) to obtain $\RateWhithin$.
At the scale of between-species variation, assuming a neutral trait evolving along the phylogeny, the mean trait value changes as a Brownian process.
This variance in mean trait value can also be normalized by the nucleotide divergence of any neutral genomic region ($d$) to obtain $\RateBetween$. [NS: -> this is repeating the mat/met, is it really needed??]
We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see methods), suggesting that traits for which this relationship is not verified are putatively under selection.
Under stabilizing selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.

[NS: this is not a result but an assumption that we have. We also already say it above. I would move it to discussion if needed]
Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate, population size and divergence time.
The relationship between trait variation between and within species can be used as a method to detect whether traits are evolving under a neutral model of evolution.
Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between-species (see fig.~\ref{fig:methods}).
This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.

>>>>

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\QstFst}{Q$_\text{ST}$--F$_\text{ST}$}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
Additionally, contrasting both trait and genetic differentiation across several populations, \QstFst methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006.
However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.

@@

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilising selection~\parencite{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
First, a trait under stabilising selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value to describe technical errors~\parencite{lynch_methods_1991, felsenstein_comparative_2008, hansen_interpreting_2012} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average as a signal of diversifying selection~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilising, and diversifying selection.
We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
For population geneticists, our study can be seen as the macro-evolutionary generalization of \QstFst methods to account for phylogenetic relationships between species.
For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the mean of variances

@@

The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the mean phenotypic trait value changes along the phylogeny, allowing estimating of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within-species, the genetic variance allows estimating of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilising selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

\subsubsection*{Bayesian estimate}

The Bayesian framework allows to obtain the posterior distribution the neutrality index ($\EstNI$) for a given trait.
Even though $\EstNI$ is estimated independently for each trait of interest in the maximum likelihood framework (previous section), here we generalize to $\Ntrait$ traits co-varying along the phylogeny using the \textit{BayesCode} software~\parencite{latrille_inferring_2021}.
Trait variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\parencite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\EstRateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is Gibbs sampled.
For each trait and each species, prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait is combined with nucleotide diversity to compute $\EstRateWhithin$ for each species, and is averaged out across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\EstRateWhithin$ and $\CovarianceMatrix$, the posterior distribution of $\EstNI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality (see fig.~\ref{fig:methods}), for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilising selection.

\subsubsection*{Applicability to empirical data}

Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\EstNI$.
It is however possible to test for diversifying selection because testing for $\EstNI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in under-estimated $\EstNI$ since $\Heritability \leq H^2$.
Contrarily, the test for stabilising selection is invalid if $\EstNI$ is under-estimated.
Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.

@@

% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
% pS = 0.0027577478571428568
% Tree length (dS) = 2.340579
% Tree length (My) = 1259.4490200000002
% Root age (My) = 98.9489413888889
% Root age (dS) = 0.18388820079995308
% nbr_sites_var = 27.577478571428568
% u = 1.3788739285714283e-05
% nbr_generations = 13336.114128321291
We tested the performance of our neutrality index ($\NI$) to detect selection on a quantitative trait using simulations.
We performed simulations under different selective regimes (neutral, stabilising, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of mutation rate (constant or fluctuating).
Simulation are individual-based and follows a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
Each individual phenotypic value is the sum of genotypic value and an environmental effect.
The environmental effect was normally distributed with variance $\VarEnv$.
We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
We scaled parameters to fit plausible values for mammals.
We used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.

The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages were both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
Additionally, the instant value of log-$\Ne$ was modeled as a independent sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.

@@

For a neutral trait, the genetic architecture, meaning the number of loci encoding the trait and the average effect of a mutation on the trait, is formally related to both within and between-species variation of the trait.
We defined the neutrality index as $\NI = \RateBetween/\RateWhithin$, which equals $1$ for a neutral trait (see methods), suggesting that traits for which this relationship is not verified are putatively under selection.
Under stabilising selection, the variation between species is depleted because the mean trait value is maintained similar between different species, which leads to $\NI < 1$.
In contrast, under diversifying selection, the variation between species is inflated because species will have potentially different trait values~\parencite{hansen_stabilizing_1997}, which leads to $\NI > 1$.
Our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between-species (see fig.~\ref{fig:methods}).
This statistic can be obtained as a maximum likelihood estimate ($\EstNI$), from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.
We devised a Bayesian estimate to obtain the posterior distribution of the neutrality index, and test for diversifying selection as $\proba [\EstNI > 1]$, and stabilising selection as $\proba [\EstNI < 1]$.

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 1.0\textwidth}
        \begin{tabular}{||l|l||c|c||c|c||}
            \hline
            Broken assumption                                       & Consequences                                       & $\EstRateWhithin$   & $\EstRateBetween$   & Test $\NI > 1$ & Test $\NI < 1$ \\ \hline \hline
            Trait encoded by few loci                        & Between-species trait variation is underestimated & --              & Underestimated & Conservative & Invalid  \\ \hline
            Sexual dimorphism                                & Within-species trait variation is overestimated   & Overestimated & -- & Conservative & Invalid  \\ \hline
            Inbreeding                                       & Nucleotide diversity ($\pi$) is underestimated    & Overestimated  & --              & Conservative & Invalid  \\ \hline
            Markers for polymorphism are negatively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
            Markers for polymorphism are positively selected & Nucleotide diversity ($\pi$) is underestimated  & Overestimated & -- & Conservative & Invalid  \\ \hline
            Markers for divergence are positively selected   & Nucleotide divergence ($d$) is overestimated & -- & Underestimated & Conservative & Invalid  \\ \hline
            Markers for polymorphism under balanced selection & Nucleotide diversity ($\pi$) is overestimated  & Underestimated & -- & Invalid & Conservative  \\ \hline
            Markers for divergence are negatively selected   & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\ \hline
            Multiple nucleotide substitutions at the same locus & Nucleotide divergence ($d$) is underestimated & -- & Overestimated & Invalid & Conservative  \\\hline \hline
        \end{tabular}
    \end{adjustbox}
    \caption{Assumptions breaks and their consequences on the estimation of within-species variation ($\EstRateWhithin$), between-species variation ($\EstRateBetween$), and on the neutrality index $\NI = \EstRateBetween/\EstRateWhithin$.
    The two last columns indicate whether the test for diversifying selection ($\NI > 1$) and for stabilising selection $\NI < 1$ are conservative or invalid due to violated assumptions.
    }
    \label{table:assumptions}
\end{table*}
